Nonparametric Clustering from Energy Statistics

### 2017-08-11 qcqp_clustering/draft/energy_clustering.tex
finished draft

\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage[none]{hyphenat}

@@

\newcommand\Energy{\mathcal{E}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\OO{{\mathcal{O}}}
%\newcommand\Zt{\widetilde{Z}}
\newcommand\Zt{Y}
%\newcommand{\Ind}[1]{\delta_{#1}}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}

@@

\begin{abstract}
Energy statistics provides a nonparametric test for equality of distributions.
It was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics. 
Energy statistics
was further generalized to probability 
distributions on arbitrary metric spaces,
and more recently a connection with reproducing kernel Hilbert spaces 
was established.
Nevertheless, although extensively used by the statistics community, it
has not been previously incorporated in machine learning problems.
In this paper, we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP).
We show the equivalence between the energy statistics clustering
formulation 
and the kernel $k$-means optimization problem (not to be confused with
kernel $k$-means algorithm). 
Therefore, our results imply a first principles derivation of 
kernel $k$-means problem
from energy statistics theory, thus bringing this important 
clustering method into a formal statistical framework.
Moreover, energy statistics is nonparametric, it usually fixes
the kernel choice,
and if prior information is available
it can be easily incorporated in the kernel construction.
We propose an iterative algorithm
to find local optimizers of the aforementioned QCQP based on the 
energy change
of moving points to different clusters. This algorithm 
is different but has the same computational cost as kernel $k$-means 
algorithm.
We compare it to
kernel $k$-means, standard $k$-means, and GMM/EM algorithms by 
providing carefully designed numerical experiments. The results
show that, in general, this algorithm outperforms
these most used clustering algorithms.
\end{abstract}

@@

\section{Introduction}

Energy statistics is based on the energy distance between probability
distributions, which provides a notion of statistical potential energy
to statistical observations, in close analogy to Newton's gravitational
potential in classical mechanics. 
It provides a nonparametric test for equality of distribution,
such that minimum energy is achieved under equality of 
distributions. 
We refer the reader 
to \cite{Szkely2013}, and references therein, for an overview.
Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, 
multi-sample tests of equality of distributions, analysis of variance
\cite{RizzoVariance}, and (nonlinear)
dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient. 
Moreover, energy statistics  was applied
to hierarchical clustering \cite{RizzoClustering} by extending Ward's
method of minimum variance.

More recently, distance covariance was 
generalized from Euclidean
spaces to metric spaces of strong negative type \cite{Lyons}.
Furthermore, a unifying framework establishing an equivalence
between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of
distributions in reproducing kernel Hilbert spaces (RKHS), was
established \cite{Sejdinovic2013}. This important work provides the missing
link
between techniques commonly used in the statistics literature, regarding
energy statistics, and techniques
commonly used in machine learning.

Given a dataset $\mathbb{X} = \{x_1,\dotsc,x_n \}$, where each
data point $x_i$ lives in a space $\mathcal{X}$, the clustering
problem consists in grouping these points into $k$ groups $\{
\C_1,\C_2,\dotsc,\C_k \}$, such that
points belonging to the same group are more ``similar'' to each other
than to points in  other groups. This intuitive notion already assumes
a ``metric'' able to measure the similarity between points. Clustering
is an unsupervised method, and one of the most important problems in machine
learning since it provides the first step towards automatically constructing
labels for previously unseen data. In practice
$k$-means is arguably the most used algorithm, and gaussian 
mixture models (GMM) through the expectation
maximization (EM) algorithm is also often used. Both are parametric, making 
strong assumptions about
the distribution of the data (isotropy and normality), and it assumes
that data points
lie in Euclidean space, $\mathcal{X} = \mathbb{R}^D$, hence
the similarity is based on Euclidean metric, 
$\| x_i - x_j \|^2 = (x_i - x_j)^\top(x_i-x_j) = 
\sum_{\ell=1}^D (x_{i,\ell} - x_{j,\ell})^2$. These algorithms provide
a good quality clustering when data is linearly separable in Euclidean space,
i.e. there is a clear resolution in the cluster centers
and points from different clusters do not overlap considerably.

To account for nonlinearly separable data, which may live
in an arbitrary non-Euclidean space, kernel methods are usually employed.
If a so-called Mercer kernel $\kk: \mathcal{X}\times \mathcal{X} \to \mathbb{R}$
is available, which is a symmetric and positive definite 
function, it guarantees that there exists a map $\varphi : \mathcal{X} \to
\mathcal{H}$ where $\mathcal{H}$ is a Hilbert space \cite{Mercer}. 
One can thus
compute similarities between data point through the inner product of
the feature space $\mathcal{H}$ and 
perform clustering in $\mathcal{H}$. The convenience of a Mercer
kernel is that the map $\varphi$ is only used implicitly, i.e. one
does not need to know $\varphi$, since the inner
of product of $\mathcal{H}$ can be computed only 
based on the kernel function $\kk(\cdot, \cdot)$ which acts on the original
space $\mathcal{X}$. 
This technique is known as the kernel trick.  
Kernel $k$-means problem 
is exactly $k$-means in the feature space
\cite{Smola,Girolami} (see also \cite{Filippone} for a survey of clustering
methods).
To cluster data that is not linearly
separable in $\mathcal{X}$,
kernel based methods exploit the nonlinear structure provided by
the kernel function, which implicitly maps data points to a usually higher
dimensional feature 
space, with the hope that they becomes separable by 
forming well-defined groups.
The choice of
kernel is crucial, and yet there is no principled method
for this. One must rely on ad hoc methods to find an appropriate kernel
which is data dependent.
As it stands, kernel $k$-means problem is an heuristic approach to clustering
in the sense that it is not derived from a statistical theory.

The main contribution of this paper is to provide a consistent formulation
to clustering based on energy statistics theory. 
The original goal is to provide a nonparametric
clustering method, since energy distance is nonparametric.
The basis of our work is the theory developed in 
\cite{Sejdinovic2013} which establishes an equivalence between energy
distance and RKHS. 
Based on this and on the multi-sample test
for equality of distributions, we show that
there is only one possible clustering optimization
problem consistent with the energy statistics formulation. 
We then show that this 
optimization problem is equivalent to a quadratically constrained quadratic
program (QCQP), which has the same trace maximization
form as kernel $k$-means,
spectral clustering, and some well-known graph partitioning problems
\cite{Dhillon2,Dhillon}. We show the equivalence between this approach
and kernel $k$-means optimization problem, implying that
the later can actually be derived from energy statistics, thus
bringing this important clustering method into a unified statistical theory.
However, it is important to note that 
energy statistics fixes a kernel to begin with.
From an algorithmic perspective, our contribution is to 
provide an iterative algorithm, which is different but has the same
complexity as kernel $k$-means algorithm. Our numerical results
provide compelling evidence that this algorithm 
is more accurate and robust than kernel $k$-means algorithm.
Moreover, these experiments 
puts in evidence the nonparametric aspect of energy
statistics based clustering, since it is able to perform accurately
on data having very different distributions, contrary to $k$-means and
GMM for instance.

Our work is organized as follows. In section~\ref{sec:background} we review
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics, leading
to a QCQP which is NP-hard.
In section~\ref{sec:twoclass} we consider a simple example in one dimension,
where we propose an algorithm which requires no initialization.
In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
and propose a new iterative algorithm to solve this QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments indicating that this algorithm outperforms kernel
$k$-means, standard $k$-means, and GMM/EM algorithms.
Our final conclusions are presented in section~\ref{sec:conclusion}.

@@

In this section we briefly review the main concepts from energy
statistics and its relation to reproducing kernel Hilbert spaces 
(RKHS) which form the basis of our work.
For more details we refer the reader
to \cite{Szkely2013} and \cite{Sejdinovic2013}.

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity \cite{Szkely2013}
\begin{equation}
\label{eq:energy}
\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
\end{equation}
called \emph{energy distance}, 
is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

The energy distance can be generalized as, for instance,
\begin{equation}
\label{eq:energy2}
\Energy_\alpha(P, Q) \equiv 
2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
\E \| Y - Y' \|^{\alpha},
\end{equation}
where $0<\alpha\le 2$. This quantity is also nonnegative,
$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$, showing that
equality to zero only requires
equality of the means, and thus $\Energy_2(P,Q)=0$ does 
not imply equality of distributions.

It is important to 
mention that \eqref{eq:energy2} can be even further generalized.
Let $X, Y \in \mathcal{X}$,  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$, and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type, there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$, which
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a legit metric.

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. Let us first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for all $x \in \mathcal{X}$.
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
all $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
there is a unique 
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
properties immediately imply that $\kk$ is symmetric and positive
definite. Indeed, notice that
$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and since
this inner product is real,
$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle = 
\langle \kkk_x, \kkk_y \rangle$, we immediately have that
the kernel is symmetric,
$\kk(y,x) = \kk(y,x)$. Moreover, for any $w \in
\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$, where
$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$,
showing that the kernel is positive definite. If $G$ is a matrix with
elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
positive semidefinite; $v^\top G \, v \ge 0$ for any vector
$v \in \mathbb{R}^n$.

The Moore-Aronszajn theorem 
\cite{Aronszajn}
establishes the converse of the above paragraph. 
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS $\Hk$ 
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical \emph{feature map}. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS as follows: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$. This is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk},
\end{equation}
which can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have 
\begin{equation}\label{eq:inner_data}
\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
\end{equation}
Thus, in practice, we can estimate the inner product between  
embedded distributions 
by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive semidefinite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type
\eqref{eq:negative_type}.
Here we have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk},
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied, we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
equivalent kernels.

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Since $\gamma_k^2(P, Q) = \| \kkk_P - \kkk_Q \|^2_{\Hk}$ we
can compute the energy distance using the inner product of $\Hk$. Moreover,
this can be computed from the data according to \eqref{eq:inner_data}.

>>>>

\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage[none]{hyphenat}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{array,booktabs}
\usepackage{xcolor}
\usepackage{makecell}

@@

\newcommand\Energy{\mathcal{E}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\OO{{\mathcal{O}}}
%\newcommand\Zt{\widetilde{Z}}
\newcommand\Zt{Y}
%\newcommand{\Ind}[1]{\delta_{#1}}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
\newcommand\e{e}
\newcommand\om{\omega}

@@

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces (RKHS)
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which is related to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
performance. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Energy statistics \cite{Szkely2013}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational
potential in classical mechanics. 
It provides a nonparametric test for equality of distribution which
is achieved under minimum energy. When probability distributions
are different, this statistical potential 
energy diverges as sample size increases.
Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, 
multi-sample tests of equality of distributions, analysis of variance
\cite{RizzoVariance}, nonlinear
dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and  
hierarchical clustering \cite{RizzoClustering} by extending Ward's
method of minimum variance. (We refer to \cite{Szkely2013} and 
references therein for an overview of energy statistics.)
Moreover, an energy statistics formulation
to clustering was already proposed in \cite{Kgroups} 
which greatly motivated this paper.

Recently, distance covariance was 
generalized from Euclidean
spaces to metric spaces of strong negative type \cite{Lyons}.
Furthermore, a unifying framework establishing an equivalence
between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of
distributions in reproducing kernel Hilbert spaces (RKHS), was
established \cite{Sejdinovic2013}. This provides the 
link between concepts used in energy statistics and concepts
commonly used in machine learning, and form the basis of our approach.

The clustering problem has a long history in machine learning.
The most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster
comes from its mean and thus it is sensitive to outliers. Moreover, $k$-means
is strongly parametric, assuming that data is isotropically and normally
distributed since it is a special case of gaussian mixture models (GMM) 
through the expectation maximization (EM) algorithm. Nevertheless,
$k$-means works very well when data is linearly separable 
in Euclidean space, with a clear resolution of cluster centers and without
significant overlap of the points belonging to different clusters.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert
space by using its inner product. This enables one to cluster way more
complicated datasets, however, the kernel choice remains the biggest challenge
since there is no principled theory behind this. The well-known
kernel $k$-means optimization problem is nothing but $k$-means in the
feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
using the mean of each cluster in the feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle 
principle behind clustering is to maximize the statistical energy, 
enforcing probability distributions associated to each cluster
to be different from one another.
We provide a precise mathematical formulation to this
statement, showing that there is only one possible optimization problem
consistent with energy statistics. Moreover, we show that this is
a quadratically constrained
quadratic program (QCQP) in the associated RKHS. Our results immediately 
establish the connection 
to kernel methods used in machine learning, 
by showing that this QCQP is equivalent
to kernel $k$-means optimization problem. 
The equivalence between kernel $k$-means, 
spectral clustering, and graph partitioning problems was already established
in \cite{Dhillon,Dhillon2}. We show how these connections arise
from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method
\cite{Hartigan} to find local solutions of the above mentioned QCQP.
This approach was also proposed in \cite{Kgroups}. In this case,
the resulting algorithm has the same
time complexity as kernel $k$-means algorithm, however, the numerical 
results provide compelling evidence that this method
is more accurate and robust than kernel $k$-means. This evidence
is also supported by general theoretical results \cite{Telgarsky,Slonin}.
Moreover, in our experiments we put in
in evidence the nonparametric aspect of energy
statistics based clustering, which provides a default kernel, showing that
it is able to perform accurately on data coming very different 
distributions, contrary to $k$-means and GMM for instance.

Our work is organized as follows. In section~\ref{sec:background} we introduce
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics leading
to a QCQP, which is NP-hard. We also show the equivalence to 
kernel $k$-means problem.
In Section~\ref{sec:weighted} we generalize these results to a weighted
version of energy statistics, which provides the connection with graph
partitioning problems and spectral clustering.
In Section~\ref{sec:twoclass} we consider a simple example in one dimension,
where we propose an algorithm which requires no initialization.
In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
and propose a new iterative algorithm based on Hartigan's method
to solve this QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments indicating that this algorithm outperforms kernel
$k$-means, standard $k$-means, and GMM/EM algorithms.
Our final conclusions are presented in section~\ref{sec:conclusion}.

@@

In this section we introduce the main concepts from energy
statistics and its relation to 
RKHS which form the basis of our work.
For more details we refer the reader
to \cite{Szkely2013} and \cite{Sejdinovic2013}.

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity 
\begin{equation}
\label{eq:energy}
\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|
\end{equation}
called \emph{energy distance} \cite{Szkely2013} 
is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

The energy distance can be generalized as, for instance,
\begin{equation}
\label{eq:energy2}
\Energy_\alpha(P, Q) \equiv 
2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
\E \| Y - Y' \|^{\alpha}
\end{equation}
where $0<\alpha\le 2$. This quantity is also nonnegative,
$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$ which shows that
equality to zero only requires
equality of the means, and thus $\Energy_2(P,Q)=0$ does 
not imply equality of distributions.

It is important to 
mention that \eqref{eq:energy2} can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a legit metric.

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. Let us first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for all $x \in \mathcal{X}$.
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
all $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
there is a unique 
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
properties immediately imply that $\kk$ is symmetric and positive
definite. Indeed, notice that
$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and by definition
$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle$, but
since the inner product is real we have 
$\langle \kkk_y, \kkk_x \rangle = \langle \kkk_x, \kkk_y \rangle$, or
equivalently 
$\kk(y,x) = \kk(x,y)$. Moreover, for any $w \in
\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$, where
$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$,
showing that the kernel is positive definite. If $G$ is a matrix with
elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
positive semidefinite, i.e. $v^\top G \, v \ge 0$ for any vector
$v \in \mathbb{R}^n$.

The Moore-Aronszajn theorem 
\cite{Aronszajn}
establishes the converse of the above paragraph. 
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS $\Hk$ 
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical \emph{feature map}. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS as follows: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$, which is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}
\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk}.
\end{equation}
This can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have 
\begin{equation}\label{eq:inner_data}
\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
\end{equation}
Thus, in practice, we can estimate the inner product between  
embedded distributions 
by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type
\eqref{eq:negative_type}.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
equivalent kernels.

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Due to \eqref{eq:mmd} we can compute the energy distance using the inner 
product of $\Hk$. Moreover, this can be estimated from a sample
according to \eqref{eq:inner_data}.
